---
title: "The Treachery of Images"
subtitle: "Exploring the Interdependence Between Graphics,<br> Statistics, and Interaction"
authors:
  - name: "Adam Bartonicek"
  - name: "Dr. Simon Urbanek"
  - name: "Dr. Paul Murrell"
affiliations: 
  - "The University of Auckland"
format: revealjs
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

## Exploring data interactively { .center .smaller .middle }
https://github.com/bartonicek/plotscape/tree/master/packages/plotscaper


```{css}

.middle {
  text-align: center
}


```

## Linked selection { }


```{r}
#| fig-align: center
#| 
library(ggplot2)
library(patchwork)

xmin <- 2
xmax <- 4.5
ymin <- 14
ymax <- 25

mtcars$x <- factor(mtcars$cyl, labels = c("A", "B", "C"))
mtcars$y <- mtcars$mpg
mtcars$zz <- factor(mtcars$wt > xmin & mtcars$wt < xmax & 
                     mtcars$mpg > ymin & mtcars$mpg < ymax)
mtcars$z <- factor(mtcars$am)

theme_set(theme_bw() +
            theme(panel.grid = element_blank(),
                  panel.border = element_blank())
          )

t <- ggplot() +
  scale_fill_manual(values = c("grey80", "indianred")) +
  scale_colour_manual(values = c("grey80", "indianred")) +
  guides(fill = "none", col = "none") +
  labs(x = NULL, y = NULL)

p1 <- t +
  geom_point(data = mtcars, aes(wt, mpg, col = zz), size = 4) +
  annotate("rect", xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
           fill = NA, col = "black", lty = "dashed")
p2 <- t + geom_bar(data = mtcars, aes(x, fill = zz))

p1 + plot_spacer() + p2 + plot_layout(widths = c(1, 0.1, 1))

```

- Aka “brushing” or “highlighting”
- Select objects and highlight the corresponding cases
- One of the most useful interactive features [@buja1996; @heer2012; @ward2015; @ware2019]

## Surprisingly tricky to implement { .center .middle }

## Some plots work better than others...

```{r}
#| fig-align: center

p3 <- t + geom_boxplot(data = mtcars, aes(x, y, fill = zz))
p2 + plot_spacer() + p3 + plot_layout(widths = c(1, 0.1, 1))
```

## Why is that? { .center .middle }

## A plot is more than just <br> geometric objects... { .center .middle }

##

![The Treachery of Images, René Magritte [-@magritte1929]](magritte2.jpg){ fig-align="center" height="550" }

##

```{r}
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
ggplot(mtcars, aes(x, y, fill = z)) +
  geom_bar(stat = "summary", fun = "mean") +
  annotate("text", 3.4, 50, 
           label = "This is not a barplot.", col = "#433A36ff",
           size = 6, family = "serif", fontface = "italic",
           vjust = 1,  hjust = 1) +
  ylim(c(0, 60)) +
  scale_fill_manual(values = c("#815A38ff", "#433A36ff")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme(
    panel.background = element_rect(fill = "#E9DEC8ff")
  )

```

## Why is that? { .center .middle }

##

```{r}
#| echo: true
#| fig-align: center

ggplot(mtcars, aes(x, y, fill = z)) +
  geom_bar(stat = "summary", fun = "mean") +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none")
```

##

```{r}
#| echo: true
#| fig-align: center
#| code-line-numbers: "2"

ggplot(mtcars, aes(x, y, fill = z)) +
  geom_bar(stat = "summary", fun = "mean") +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none")
```

##

```{r}
#| echo: true
#| fig-align: center
#| code-line-numbers: "2"

ggplot(mtcars, aes(x, y, fill = z)) +
  geom_bar(stat = "summary", fun = "mean", position = "stack") +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none")
```

## The height of the stacked bars is not a valid summary statistic... { .center .middle }

## { .center }

>"Stacking is useful when the sum of the amounts represented by the individual stacked bars is in itself a meaningful amount" [@wilke2019, p. 52].

## { .center }

>"Because this gives the visual impression of one element that is the sum of several others, it is very important that if the element’s size is used to display a statistic, then that statistic must be summable. Stacking bars that represent counts, sums, or percentages are fine, but a stacked bar chart where bars show average values is generally meaningless." [@wills2011, p. 112].

## Stacking some statistics is okay, stacking others is not { .center .middle }

## Can we make this idea precise? Are other statistics "stackable"? { .center .middle }

## { .center .middle }

> The ***sum*** of group ***sum***s is equal to the ***sum*** on the whole

## { .center .middle }

> The ***foo*** of group ***foo***s is equal to the ***foo*** on the whole

## Other statistics behave this way too... { .center .middle }

##

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

library(ggplot2)
library(patchwork)

stack <- function(values, fn, initial) {
  values[is.na(values)] <- initial
  for (i in seq_along(values)) values[i] <- fn(values[i - 1], values[i])
  values
}

group <- factor(rep(c("A", "B"), each = 2))
subgroup <- factor(rep(1:2, 2))
value <- c(NA, 1.5, 1.1, 1.6)

sum_value <- rev(unlist(tapply(value, group, function(x) stack(x, sum, 0))))
prod_value <- rev(unlist(tapply(value, group, function(x) stack(x, prod, 1))))
max_value <- rev(unlist(tapply(value, group, function(x) stack(x, max, 0))))

or_na <- function(x) ifelse(is.na(x), "NA", x)
show_value <- or_na(value[c(3, 4, 1, 2)])

d1 <- data.frame(group, subgroup, value = value[4:1], 
                 sum_value, prod_value, max_value)

format_label <- function(value, symbol) {
  ifelse(is.na(value), value, paste(symbol, value))
}

p0 <- ggplot(d1) +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, hjust = 0.1, vjust = -3),
    panel.border = element_blank(),
)

p2 <- p0 +
  geom_bar(mapping = aes(group, sum_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_label(aes(group, sum_value, label = format_label(value, "+")),
             size = 3) +
  scale_y_continuous(limits = c(0, 3)) +
  labs(title = "Sum")

p3 <- p0 +
  geom_bar(mapping = aes(group, prod_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 0, ymax = 1), 
            fill = "white") + # Use to mask the bar starting from 0
  geom_label(aes(group, prod_value, label = format_label(value, "x")),
             size = 3) +
  coord_cartesian(ylim = c(1, 1.8)) + 
  labs(title = "Product")

p4 <- p0 +
  geom_bar(mapping = aes(group, max_value, fill = subgroup), 
           stat = "identity", position = "identity") +
  geom_label(aes(group, max_value, label = format_label(value, "max")),
             size = 3) +
  scale_y_continuous(limits = c(0, 1.75)) +
  labs(title = "Maximum")

p2 + p3 + p4 + plot_layout(nrow = 1)


```

## We need some math... { .center .middle }

## Monoids { .center }

- A monoid is a tuple $(M, \otimes, e)$ consisting of of:
  - A set of objects $M$
  - A binary operation $\otimes: M \times M \to M$
  - A neutral object $e \in M$
- Subject to two rules...

## Monoids: Rules { .center }

- Unitality: $x \otimes e = e \otimes x = x$
- Associativity: $x \otimes (y \otimes z) = (x \otimes y) \otimes z$

## Example: Sums { .center .middle }

$$1 + 0 = 0 + 1 = 1$$
$$1  + (2 + 3) = (1 + 2) + 3$$

## Example: Products { .center .middle }

$$1 \times 2 = 2 \times 1 = 2$$
$$2 \times (3 \times 4) = (2 \times 3) \times 4$$

## Example: Maximum { .center .middle }

$$\max(x, -\infty) = \max(-\infty, x) = x$$
$$\max(x, \max(y, z)) = \max(\max(x, y), z)$$

## Counterexample: Exponentiation { .center .middle }

$$(x)^1 = x \qquad \text{but} \qquad 1^x \neq x$$
$$(x^y)^z \neq x^{(y^z)}$$


## Counterexample: Averages { .center .middle }

$$\text{mean}(x, ?) = x$$
$$\text{mean}(x, \text{mean}(y, z)) \neq \text{mean}(\text{mean}(x, y), z)$$

## Monoids preserve set union { .center }

- Suppose we have some subsets of the data $A, B \subseteq D$, and a monoidal summary $F$ such that e.g.:

$$F(A) = a_1 \otimes a_2 \otimes \ldots \otimes a_n$$

- Combining summaries is the same as summarizing union:

$$\begin{aligned}
F(A) \otimes F(B) &= (a_1 \otimes a_2 \otimes \ldots a_n) \otimes (b_1 \otimes b_2 \otimes \ldots b_m) \\
&= a_1 \otimes a_2 \otimes \ldots a_n \otimes b_1 \otimes b_2 \otimes \ldots b_m \\
&= F(A \cup B)
\end{aligned}$$

## This diagram commutes...

```{tikz}
#| fig-align: center

\usetikzlibrary{cd}

\begin{tikzcd}
	& A \\
	{A \cup B} && {A \cup B \cup C} \\
	& {F(A)} \\
	{F(A \cup B)} && {F(A \cup B \cup C)}
	\arrow["{- \cup B}"', from=1-2, to=2-1]
	\arrow["{- \cup (B \cup C)}", from=1-2, to=2-3]
	\arrow[dashed, from=1-2, to=3-2]
	\arrow["{- \cup C}"', near start, from=2-1, to=2-3]
	\arrow[dashed, from=2-1, to=4-1]
	\arrow[dashed, from=2-3, to=4-3]
	\arrow["{- \otimes F(B)}"', near start, from=3-2, to=4-1]
	\arrow["{- \otimes F(B \cup C)}", near start, from=3-2, to=4-3]
	\arrow["{- \otimes F(C)}"', from=4-1, to=4-3]
\end{tikzcd}
```

## We can compare nested subsets! { .center .middle }

$F(A)$ vs. $F(A \cup B)$

## References